---
title: "Supervised Learning Tools for Deriving Biomarkers based on Single-Cell Data"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: 
  - name: Tingting Zhan
    orcid: 0000-0001-9971-4844
    email: tingtingzhan@gmail.com
    affiliations: 
      - ref: tjuh
  - name: Inna Chervoneva
    orcid: 0000-0002-9104-4505
    email: Inna.Chervoneva@jefferson.edu
    affiliations: 
      - ref: tjuh
affiliations:
  - id: tjuh
    name: Thomas Jefferson University & Hospitals
    address: 130 South 9th Street
    city: Philadelphia
    state: PA
    postal-code: 19107
format: 
  html:
    page-layout: full
    html-math-method: katex
    link-external-icon: true
    link-external-newwindow: true
number-sections: true
toc: true
toc-location: left
toc-depth: 4
toc-title: ''
editor: source
bibliography: hypergam.bib
knitr:
  opts_chunk: 
    collapse: true
    comment: "#>" 
vignette: >
  %\VignetteIndexEntry{applications}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The complete vignette of R package **`hyper.gam`** (v`r packageVersion('hyper.gam')`) 

> <https://rpubs.com/tingtingzhan/hyper_gam>

documents the derivation of single index predictors of scalar outcomes based on spatial and non-spatial single-cell imaging data.

The complete vignette exceeds the file size limit allowed on [CRAN](https://cran.r-project.org/package=hyper.gam).  



# Prerequisite

## Environment

Package **`hyper.gam`** (v`r packageVersion('hyper.gam')`) requires R version 4.5.0 (released 2025-04-11) or higher ([macOS](https://cran.r-project.org/bin/macosx/), [Windows](https://cran.r-project.org/bin/windows/base/), [Linux](https://cran.r-project.org/bin/linux/)).

An Integrated Development Environment (IDE), e.g., [RStudio](https://posit.co/download/rstudio-desktop/) [@RStudio] or [Positron](https://positron.posit.co/download.html), is not required, but highly recommended.


```{r}
#| code-fold: true
#| code-summary: "Environment on author's computer"
#| label: author-env
Sys.info()[c('sysname', 'release', 'machine')]
R.version
```


## Enhancement & Dependency

Package **`hyper.gam`** (v`r packageVersion('hyper.gam')`) `Enhances` package **`mgcv`** [@mgcv, v`r packageVersion('mgcv')`]. Details are provided in the [complete vignette](https://rpubs.com/tingtingzhan/hyper_gam).


The dependencies of package **`hyper.gam`** are detailed in the [complete vignette](https://rpubs.com/tingtingzhan/hyper_gam). 


Package **`hyper.gam`** requires the *development versions* of the **`spatstat.*`** family of packages [@spatstat15; @spatstat05]. Installation instructions are provided in the [complete vignette](https://rpubs.com/tingtingzhan/groupedHyperframe) of package **`groupedHyperframe`** [@groupedHyperframe, v`r packageVersion('groupedHyperframe')`], Section 4.1. 



## Installation

Packages **`hyper.gam`** (v`r packageVersion('hyper.gam')`) and **`groupedHyperframe`** (v`r packageVersion('groupedHyperframe')`) can be installed using the following command.

```{r}
#| eval: false
utils::install.packages('groupedHyperframe')
utils::install.packages('hyper.gam')
```

## Getting Started

Examples in the [complete vignette](https://rpubs.com/tingtingzhan/hyper_gam) require that the `search` path has

```{r}
#| message: false
library(groupedHyperframe)
library(hyper.gam)
library(survival)
```


```{r}
#| echo: false
op = par(no.readonly = TRUE)
#options(mc.cores = 1L) # for CRAN submission
```

## Acknowledgement

This work is supported by National Institutes of Health, U.S. Department of Health and Human Services grants

-   R01CA222847 ([I. Chervoneva](https://orcid.org/0000-0002-9104-4505), [T. Zhan](https://orcid.org/0000-0001-9971-4844), and [H. Rui](https://orcid.org/0000-0002-8778-261X))
-   R01CA253977 (H. Rui and I. Chervoneva).

The authors thank

-   [Erjia Cui](https://orcid.org/0000-0003-3576-2892)'s contribution to function `hyper_gam()`.



# References {.appendix}

::: {#refs}
:::
