---
title: "Generalized Additive Model with Hyper-Column Predictor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

This vignette of package **`hyper.gam`** ([Github](https://github.com/tingtingzhan/hyper.gam),
[RPubs](https://rpubs.com/tingtingzhan/hyper_gam)) documents ...

## Note to Users

Examples in this vignette require that the `search` path has

```{r setup}
library(hyper.gam)
library(groupedHyperframe)
library(survival)
```

## Terms and Abbreviations

```{r echo = FALSE, results = 'asis'}
c(
#  '`abs`', 'Absolute value', '`base::abs`',
  '`coxph`', 'Cox model', '`survival::coxph`',
#  '`factor`', 'Factor, or categorical variable', '`base::factor`',
#  '`function`', 'R function', '``base::`function` ``',
#  '`groupedHyperframe`', 'Grouped hyper data frame', ' https://CRAN.R-project.org/package=groupedHyperframe',
#  '`head`', 'First parts of an object', 'Generic `utils::head`; Dispatch `utils:::head.default`',
  '`htmlwidget`', 'HTML Widgets', 'R package [**`htmlwidgets`**](https://CRAN.R-project.org/package=htmlwidgets), [**`plotly`**](https://CRAN.R-project.org/package=plotly)',
#  '`levels`', 'Levels of a `factor`', '`base::levels`',
#  '`listof`', 'List of objects (of same class)', '`stats::listof`',
#  '', 'Logistic regression model', '`stats::glm(., family = binomial(\'logit\'))`',
#  '`matrix`', 'Matrix', '`base::matrix`',
  '`PFS`', 'Progression/recurrence free survival', 'https://en.wikipedia.org/wiki/Progression-free_survival'
#  '`quantile`', 'Quantile', '`stats::quantile`',
#  '`rpart`', 'Recursive partitioning and regression trees', '`rpart::rpart`',
#  '`S3`', '`R`\'s simplest object oriented system', 'https://adv-r.hadley.nz/s3.html',
#  '`Surv`', 'Survival object', '`survival::Surv`',
#  '`update`', 'Update and re-fit a model call', 'Generic `stats::update`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable()
```



## Acknowledgement

This work supported by NCI R01CA222847 ([I. Chervoneva](https://orcid.org/0000-0002-9104-4505), [T. Zhan](https://orcid.org/0000-0001-9971-4844), and [H. Rui](https://orcid.org/0000-0002-8778-261X)) and R01CA253977 (H. Rui and I. Chervoneva). 




# Data Preparation

Data preparation

```{r}
data(Ki67, package = 'groupedHyperframe')
Ki67
```

```{r}
s = Ki67 |>
  aggregate_quantile(by = ~ patientID, probs = seq.int(from = .01, to = .99, by = .01))
s |> head()
```

Users are encouraged to learn more about `groupedHyperframe` class and function `aggregate_quantile()` from package **`groupedHyperframe`** [vignettes](https://rpubs.com/tingtingzhan/groupedHyperframe).


```{r}
m0 = hyper_gam(PFS ~ logKi67.quantile, data = s)
```




Function `integrandSurface()` creates .. an `htmlwidget` object using package **`plotly`**.
```{r}
integrandSurface(m0)
#integrandSurface(m0, newdata = s_1)
```


```{r}
m1 = hyper_gam(PFS ~ logKi67.quantile, data = s, nonlinear = TRUE)
```

```{r}
integrandSurface(m1)
#integrandSurface(m1, newdata = s_1)
```

```{r}
integrandSurface(m0, m1)
```


```{r}
#fr = sign_adjust(m0)
#predict(m0, newdata = s_1)
m0 |> predict_k_fold.hyper_gam(k = 10L)
persp(m0)
```




```{r}
set.seed(145); pred1 = m1 |> predict_k_fold.hyper_gam(k = 10L)

pred1 |> with.default(expr = boxplot(est ~ fold))
pred1 |> with.default(expr = boxplot(est.k ~ fold)) 
pred1 |> with.default(expr = boxplot(est.global ~ fold)) 
table(pred1$signadj)

s. = data.frame(s, pred1[c('est', 'est.k', 'est.global')])

coxph(PFS ~ est, data = s.) |> summary()
coxph(PFS ~ est.k, data = s.) |> summary()
coxph(PFS ~ est.global, data = s.) |> summary()
```
